Sampling Strategies in Decisions from Experience

### 2022-02-04 manuscript.Rmd
Update Introduction

The human mind is a cognitive system that operates on inputs from an environment or memory. 
In the domain of decision making, such inputs provide information about the choice alternatives among which people can choose. 
Given that each choice alternative is an element $i$ of the set $I$, the basic structure of decision making may be modeled by the inclusion map
$$
\begin{aligned}
  \iota : I' &\to I \\
  i &\mapsto i \quad \text{for all} \ i \in I' 
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, pp. 14-15], indicating that decision making is an information processing problem which starts from an input set $I$, including all choice alternatives, and stops at a proper subset $I' \subset I$, including only the chosen alternatives.
Indeed, irrespective of their exact mathematical specification or computational implementation, decision models generally take as inputs information about a set of choice alternatives to return a subset thereof as outputs [@heOntologyDecisionModels2020].

However, information about the choice alternatives may come in many forms.
Rarely in their daily life, people make *decisions from description*, since in such a case all information about the choice alternatives is available through symbolic descriptions of their *properties*. 
Yet, rather often, people make *decisions from experience* [@hertwigDecisionsExperienceEffect2004], since

# Header

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from particular decisions, e.g., the choice between job offers, political parties to vote for, travel destinations, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (or: gambles), it studies a case which is counterfactual in that it omits the many particularities of each decision and the differences between them, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983; see also @knightRiskUncertaintyProfit1921, for a discussion of the reasonableness of such abstractions]. 
These fundamental properties are considered to be the possible outcomes of a choice alternative---hereafter denoted by the Greek letter $\omega$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p(\omega)$.

Now, let a prospect be a *probability space* denoted by the tuple $(\Omega, \mathcal{F}, P)$. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ denotes the sample space, containing a finite set of possible outcomes:
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \mapsto [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$, also referred to as *events*, a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \mapsto [0,1]$, which is used to construct the probability measure $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$ [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.2].]

The choices people actually make between such prospects are often described in terms of deviations from expected value (EV) maximization, where the EV is given by 
$$
EV = \sum_i^n \omega_i \times P(X = \omega_i)
\; ,
\tag{4}
$$ 
with $X$ being a random variable defined on the respective probability space.^[$X$ is measurable function $X: \Omega \to \Omega'$ [see @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.3] with $\omega \mapsto \omega$ for all $\omega \in \Omega$.]
That is, following the principle of EV maximization, one would first calculate the prospects' EV according to (3) and thereafter chose the prospect with the largest EV.
To *describe* how peoples' choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data with a combination of two parameterized functions that may replace the objective outcomes and probabilities in (3) by nonlinear transformations thereof.^[Note that CPT or other originally deterministic decision models may be amended by a stochastic choice function [see, e.g., @rieskampProbabilisticNaturePreferential2008].] 
That is, the models of CPT fit choice data by maximizing the value 
$$
V = \sum_i^n v(\omega_i) \times \pi_i
\; ,
\tag{5}
$$ 
with subjective values $v(\omega)$ derived from a *value function* that may transform the objective outcomes, and decision weights $\pi$ derived from the objective probabilities' cumulative distribution function that may be transformed by a *probability weighting function* (see below). 
Now, since the probability weighting function may redistribute the entire probability mass of $\sum_i^n p(\omega_i) = 1$ across outcomes, the decision weights 
$$
\pi =
  \begin{cases}
     w(P(X \geq \omega_i)) - w(P(X > \omega_i)) \quad \text{for} \ 1, \\
     w(P(X \leq \omega_i)) - w(P(X < \omega_i)) \quad \text{else}, 
  \end{cases}
\tag{5}
$$
may 
people's deviations from EV maximization is often described in terms of the probability weighting pattern.

>>>>

The human mind is a cognitive system that operates on inputs from an environment or memory. 
In the domain of decision making, such inputs provide information about the choice alternatives among which people can choose and may therefore be organized in a set $I$ with each element $i \in I$ denoting a choice alternative. 
That is, each element $i \in I$ is assigned some information from the inputs and this information may be used by the mind to learn about the respective choice alternative.
In turn, decision making can be considered an information processing problem which starts from the set $I$, including all choice alternatives, and stops at a proper subset $I' \subset I$, including only the chosen alternatives.
Indeed, irrespective of their exact mathematical specification or computational implementation, decision models generally take as inputs information about a set of choice alternatives to return a subset thereof as outputs [@heOntologyDecisionModels2020].

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, travel destinations, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental *properties* that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, for a discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered to be the possible outcomes of a choice alternative---hereafter denoted by the Greek letter $\omega$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p(\omega)$.
That is, any choice alternative that can be fully described by its outcomes and their probabilities is a prospect. 
For illustration, take the choice alternatives in the following example with monetary outcomes:

\noindent
**Example 1.** 
*A1 gives \$1 with 100%.*
*A2 gives \$10 with 90% and \$0 with 10%.*

Now, let the exact axiomatic definition of a prospect be that of a *probability space* denoted by the tuple $(\Omega, \mathcal{F}, P)$. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ denotes the sample space, containing a finite set of possible outcomes:
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{1}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$, with each element $A \in P(\Omega)$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \mapsto [0,1]
\; ,
\tag{2}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \mapsto [0,1]$, which is used to construct the probability measure $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]
Given a choice between 1, ..., k prospects, the inputs from the environment or memory should therefore provide information about the set
$$
I = \{(\Omega_1, \mathcal{F}_1, P_1), ... ,(\Omega_k, \mathcal{F}_k, P_k)\}
\; .
\tag{3}
$$

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes and their probabilities (as in Example 1).
In such cases, people would make *decisions from description* (DfD), since the information from which the mind learns about the choice alternatives---the probability spaces in (3)---are complete descriptions thereof.
Yet, rather often, people make *decisions from experience* [DFE, @hertwigDecisionsExperienceEffect2004], where the mind can learn about the choice alternatives only from a record of past experiences with them, which necessarily provides only incomplete and probable information.
That is, people may have experienced the possible outcomes of a choice alternative when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular choice alternative is denoted a *sample*. 
The mind may then infer from the relative frequency with which an outcome occurs in such a sample of past decisions its probability of occurrence for a future decision. 
Of course, all other things being equal, the accuracy of this inference depends on the size of the sample.
<!--For illustration, take the sampled outcomes in the following example, where alternative A2 from Example 1 was chosen ten times without prior knowledge about the possible outcomes of this alternative and their probabilities:

\noindent
**Example 2.** *A2 gave: \$0, \$0, \$10, \$0, \$0, \$0, \$0, \$10, \$0, \$0*

\noindent
One may now make an inference from the sample about the possible outcomes of choice alternative A2 and their probabilities. 
However, this conclusion is not certain, but only probable, since the relative frequencies of the results will almost surely change as the sample size increases.-->

To provide a mathematical definition, this paper treats a sample as a collection of independent and identically distributed (i.i.d.) random variables $X$, each of which is a measurable function defined on the same probability space:^[Note that this paper considers the counterfactual case where samples are random, although it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. To account for non-random sampling, e.g., in the form of sampling biases [e.g., @fiedlerBewareSamplesCognitiveecological2000] or dynamic information structures [e.g., @cohenEffectPerceivedPatterns2021; @plonskyRelianceSmallSamples2015], the i.i.d-assumption would need to be dropped, which is out of the scope of this paper.] 
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F}') \\
  \omega &\mapsto \omega  \quad \text{for all} \ \omega \in \Omega
\end{aligned}
\tag{4}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21; see also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.3, for a more accessible and comprehensive treatment]. 
Accordingly, the inputs to DfE 
$$
S = \{X_{1i}, ... , X_{ki}\}
\; ,
\tag{5}
$$
where $i \in \{1, ..., n\}$.

The choices people actually make between prospects---be that on the basis of ---are often described in terms of deviations from expected value (EV) maximization, where the EV is given by 
$$
EV = \sum_i^n \omega_i \times P(X = \omega_i)
\; .
\tag{6}
$$ 
That is, following the principle of EV maximization, one would first calculate the prospects' EV according to (6) and thereafter chose the prospect with the largest EV.
To *describe* how peoples' choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data with a combination of two parameterized functions that may replace the objective outcomes and probabilities in (3) by nonlinear transformations thereof.^[Note that CPT or other originally deterministic decision models may be amended by a stochastic choice function [see, e.g., @rieskampProbabilisticNaturePreferential2008].] 
That is, the models of CPT fit choice data by maximizing the value 
$$
V = \sum_i^n v(\omega_i) \times \pi_i
\; ,
\tag{5}
$$ 
with subjective values $v(\omega)$ derived from a *value function* that may transform the objective outcomes, and decision weights $\pi$ derived from the objective probabilities' cumulative distribution function that may be transformed by a *probability weighting function* (see below). 
Now, since the probability weighting function may redistribute the entire probability mass of $\sum_i^n p(\omega_i) = 1$ across outcomes, the decision weights 
$$
\pi =
  \begin{cases}
     w(P(X \geq \omega_i)) - w(P(X > \omega_i)) \quad \text{for} \ 1, \\
     w(P(X \leq \omega_i)) - w(P(X < \omega_i)) \quad \text{else}, 
  \end{cases}
\tag{5}
$$
may 
people's deviations from EV maximization is often described in terms of the probability weighting pattern.
